Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-07-06 paper/paper.tex
Finished proofreading

In the previous section, we have established a computable pushdown control-flow
analysis through refunctionalization and caching. In this section, we revisit
the pushdown control-flow problem and examine what we have done to overcome it. \\

\textit{The Problem with Return-flows.}
Pushdown control-flow is a property in the analysis that precisely models the run-time
call structure of the analyzed program. A pushdown control-flow analysis provides an 
as-exact-as-run-time return-flow when analyzing a program, but traditional 
control-flow analysis collapses the state space into a finite space and thus 
causes imprecise stack modeling.

@@

In $k$-CFA algorithm or using the abstract abstract machine shown in Section~\ref{aam},
the call sites @(id 2)@ and @(id 1)@ share the same return flow, so the invocation
of @(id 2)@ returns to both call-sites @[x (id 1)]@ and @[y (id 2)]@.  Therefore,
the returned value @2@ for variable @y@ is also propagated to variable @x@, causing
imprecise analysis results to arise. This return-flow merging is inevitable, even when
increasing the context-sensitivity. Even worse, if we use the monovariant analysis (i.e.,
$0$-CFA), the analysis result would be such that @x@ and @y@ point to the value set @{1, 2}@.
Because the algorithm does not distinguish incoming values from different call sites 
for @z@ (the argument of @id@). Under $1$-CFA, the algorithm is able to distinguish that variable
@z@ has two different values at two call sites, so variable @y@ would not be polluted by
@1@. However, variable @x@ still points to the value set @{1, 2}@, as the two call-sites 
still share the same continuation and $1$-CFA does not increase the ability to separate the 
return-flows. \\

@@

In fact, what we started with is an unbounded-stack A\textsuperscript{2}M which already precisely
matches the calls and returns. Refunctionalization makes the interpreter a ``big-step''
semantic artifact; the consequence is that we have no place to explicitly save the context
information in the store or on the stack. 
But refunctionalization not only forces the nondeterministic control flow to be higher-order,
but also drives us to use a different caching algorithm.

Indeed, the caching algorithm plays an important role to guarantee the analysis will
terminate eventually.
An interesting implication of this is if we apply the caching algorithm with some 
necessary changes to small-step abstract machines with an unbounded stack, we are 
also able to establish a computable and precise call/return match.

@@

\begin{itemize}
  \item By replacing the continuations with monads \cite{filinski1994representing}, 
    we obtain the abstract definitional interpreters in monadic style which are similar to what 
    \citet{darais2017abstracting} describe.
  \item By representing the continuation-passing style with delimited control operators
    such as @shift@ and @reset@, we derive a new form of abstract interpreters in direct-style.
    This transformation is the left-inverse of the CPS transform \cite{DBLP:journals/scp/Danvy94, DBLP:conf/lfp/DanvyL92}.
  \item In fact, we may also just use the same caching algorithm but with side effects such as
    assignments and mutations to update the cache, and then achieve the same definitional 
    interpreter with pushdown control-flows. \Citet{friedman-ai} in a tutorial on abstract 
    interpreters use this style to model caches and updates.
    In this case, nondeterminism can be easily handled via @for@ comprehensions. 
\end{itemize}

These coincidences should not be a surprise since the literature is rich
in showing the correspondence between monads and continuation-passing style as well as
delimited control \cite{Danvy:1990:AC:91556.91622, wadler1992essence,
danvy1992representing, moggi1991notions}.

@@

We first transform the function @aeval@ to direct-style by removing the additional 
continuation argument @continue@ and sequentializing the expressions in @aeval@.
To enable using delimited control operators inside of @aeval@, we also add an 
annotation \verb|@cps[Ans]| on the return type, which tells the compiler to CPS
transform this function.
Back to direct-style on the case where @e@ appears as an atomic expression is straightforward: 
we simply unwrap the application of @continue@. Now the abstract evaluator @aeval@ looks as follows:

@@

The case where @e@ is a @let@ expression is not that obvious. As previously identified, 
function @nd@ works on an unknown number of nondeterministic choices, and it takes a 
function @k@ that works for every element in the collection. Here we present another
function @choices@ that uses delimited control operator @shift@ to capture that ``@k@''
when the program is written in direct-style and pass it to @nd@.

@@

The function @choices@ takes two arguments: a set of elements of type @T@, and an initial cache.
The return type of @choices@ is @(T, Cache)@ with an annotation \verb|@cps[Ans]|, which 
denotes that @choices@ will iteratively return an element of type @T@ from the collection
along with the latest cache once @f@ is invoked, but the final returned value of the 
function is type @Ans@.
The function @f@ is the delimited continuation captured by the @shift@ operator, which
is formed by the computation will be executed after the @choices@'s call-site.
Once the function @f@ (which is the function @k@ in @nd@) is invoked and returned, 
it returns to its call-site where @nd@ will be invoked on the remaining elements of the collection.
Now we can use this function to sequentialize expressions in the @let@ case.

@@

For every nondeterministic fork-point in the abstract interpretation, we can simply call 
@choices@ on the set of possible choices and write the program sequentially as concrete 
interpreters, and the subsequent statements and expressions after the call-site become 
the delimited continuation @f@. We have two applications of @choices@, which correspond 
to the two applications of @nd@ in the refunctionalized form in the last section.
The function @choices@ also returns the latest cache to its left-hand side when invoked.
The cache will be still accumulated to the following calls of @choices@ or @aeval@, 
given the fact that the abstract interpreter should always use the latest cache.

The function @analyze@ is also changed by calling the @reset@ operator around the function 
application of @aeval@ to set the delimiter.

@@

Now we have arrived at the end of this series of transformations. 
The inverse transformation of the last step is basically CPS transforming the function @aeval@ and
explicitly using function @nd@ instead of function @choices@. Then, these continuations must be 
manifestly embedded at the call-site of @nd@, instead of implicitly provided by the @shift@ operator.

@@

In this Functional Pearl, we bridge the gap between small-step abstract abstract machines
and big-step abstract definitional interpreters by applying a series of syntactic 
transformations to transform the first into the latter. Among these transformations, 
linearization turns a work-list into an additional layer of continuations; refunctionalization 
converts the first-order data types representing continuations to higher-order functions; 
and finally the left-inverse of the CPS transformation converts the CPS abstract interpreter 
into a direct-style abstract interpreter with delimited control operators, which looks almost 
identical to the corresponding concrete interpreter.

This sequence of transformation demonstrates that a functional correspondence exists 
not only between concrete semantic artifacts but also between abstract semantic artifacts. 
An interesting open question would be whether there also exist correspondences between 
static analyses that are formalized 
in different denotational and operational styles.

>>>>

In the previous section, we established a computable pushdown control-flow
analysis through refunctionalization and caching. In this section, we revisit
the pushdown control-flow problem and examine what we have done to overcome it. \\

\textit{The Problem with Return-flows.}
Pushdown control-flow is a property in the analysis that precisely models the run-time
call structure of the analyzed program. A pushdown control-flow analysis provides an
as-exact-as-runtime return-flow when analyzing a program, but traditional
control-flow analysis collapses the state space into a finite space and
thus causes imprecise stack modeling.

@@

In the $k$-CFA algorithm, or in using the abstract abstract machine shown in Section~\ref{aam},
the call sites @(id 2)@ and @(id 1)@ share the same return flow, so the invocation
of @(id 2)@ returns to both call-sites @[x (id 1)]@ and @[y (id 2)]@.  Therefore,
the returned value @2@ for variable @y@ is also propagated to variable @x@, causing
imprecise analysis results to arise. This return-flow merging is inevitable, even when
increasing the context-sensitivity. Even worse, if we use the monovariant analysis (i.e.,
$0$-CFA), the analysis result would be such that @x@ and @y@ point to the value set @{1, 2}@.
Because the algorithm does not distinguish incoming values from different call sites
for @z@ (the argument of @id@). Under $1$-CFA, the algorithm is able to distinguish that variable
@z@ has two different values at two call sites, so variable @y@ would not be polluted by
@1@. However, variable @x@ still points to the value set @{1, 2}@, as the two call-sites
still share the same continuation and $1$-CFA does not increase the ability to separate the
return-flows. \\

@@

In fact, we started with an unbounded-stack A\textsuperscript{2}M which already precisely
matches the calls and returns. Refunctionalization makes the interpreter a ``big-step''
semantic artifact; the consequence is that we have no place to explicitly save the context
information in the store or on the stack.
However, refunctionalization not only forces the nondeterministic control flow to be higher-order,
but also drives us to use a different caching algorithm.

Indeed, the caching algorithm plays an important role in guaranteeing
the analysis will eventually terminate.
An interesting implication of this is if we apply the caching algorithm with some
necessary changes to small-step abstract machines with an unbounded stack, we are
also able to establish a computable and precise call/return match.

@@

\begin{itemize}
  \item By replacing the continuations with monads \cite{filinski1994representing},
    we obtain the abstract definitional interpreters in monadic style which are similar to what
    \citet{darais2017abstracting} describe.
  \item By representing the continuation-passing style with delimited control operators
    such as @shift@ and @reset@, we derive a new form of abstract interpreters in direct-style.
    This transformation is the left-inverse of the CPS transform \cite{DBLP:journals/scp/Danvy94, DBLP:conf/lfp/DanvyL92}.
  \item In fact, we may also just use the same caching algorithm but with side effects such as
    assignments and mutations to update the cache, and then achieve the same definitional
    interpreter with pushdown control-flows. \Citet{friedman-ai}, in a tutorial on abstract
    interpreters, use this style to model caches and updates.
    In this case, nondeterminism can be easily handled via @for@ comprehensions.
\end{itemize}

These coincidences should not be a surprise, since the literature is rich
in showing the correspondence between monads and continuation-passing style as well as
delimited control \cite{Danvy:1990:AC:91556.91622, wadler1992essence,
danvy1992representing, moggi1991notions}.

@@

We first transform the function @aeval@ to direct-style by removing the additional
continuation argument @continue@ and sequentializing the expressions in @aeval@.
To enable using delimited control operators inside of @aeval@, we also add an
annotation \verb|@cps[Ans]| on the return type, which tells the compiler to CPS
transform this function.
Returning to direct-style in the case where @e@ appears as an atomic expression is straightforward:
we simply unwrap the application of @continue@. The abstract evaluator @aeval@ now looks as follows:

@@

The case where @e@ is a @let@ expression is less straightforward. As
previously identified, @nd@ works on an unknown number of
nondeterministic choices, and it takes a
function @k@ that works for every element in the collection. Here we present another
function @choices@ that uses delimited control operator @shift@ to capture that @k@
when the program is written in direct-style and pass it to @nd@.

@@

The function @choices@ takes two arguments: a set of elements of type @T@, and an initial cache.
The return type of @choices@ is @(T, Cache)@ with an annotation \verb|@cps[Ans]|, which
denotes that @choices@ will iteratively return an element of type @T@ from the collection
along with the latest cache once @f@ is invoked, but the final returned value of the
function is type @Ans@.
The function @f@ is the delimited continuation captured by the @shift@ operator (formed by the computation), and will be executed after the @choices@'s call-site.
Once the function @f@ (which is the function @k@ in @nd@) is invoked and returned,
it returns to its call site where @nd@ will be invoked on the remaining elements of the collection.
Now we can use this function to sequentialize expressions in the @let@ case:

@@

For every nondeterministic fork-point in the abstract interpretation, we can simply call
@choices@ on the set of possible choices and write the program sequentially as concrete
interpreters, and the subsequent statements and expressions after the call site become
the delimited continuation @f@. We have two applications of @choices@, which correspond
to the two applications of @nd@ in the refunctionalized form in the previous section.
@choices@ also returns the latest cache to its left-hand side when invoked.
The cache will continue to accumulate for the following calls of @choices@ or @aeval@,
given the fact that the abstract interpreter should always use the latest cache.

@analyze@ is also changed by calling the @reset@ operator around the
function application of @aeval@ to set the delimiter.

@@

We have finally arrived at the end of this series of transformations.
The inverse transformation of the last step is essentially that of performing a CPS transformation on the function @aeval@ and
explicitly using function @nd@ instead of function @choices@. Then, these continuations must be
manifestly embedded at the call site of @nd@, instead of implicitly provided by the @shift@ operator.

@@

In this Functional Pearl, we bridge the gap between small-step abstract abstract machines
and big-step abstract definitional interpreters by applying a series of syntactic
transformations to transform the former into the latter. Among these transformations,
linearization turns a work-list into an additional layer of continuations, refunctionalization
converts the first-order data types representing continuations to higher-order functions,
and finally, the left-inverse of the CPS transformation converts the CPS abstract interpreter
into a direct-style abstract interpreter with delimited control operators, which looks almost
identical to the corresponding concrete interpreter.

This sequence of transformation demonstrates that a functional correspondence exists
not only between concrete semantic artifacts, but also between abstract semantic artifacts.
An interesting open question remains of whether there also exist
correspondences between static analyses that are formalized in different
denotational and operational styles.
